\documentclass[11pt]{report}
\usepackage{multicol,lipsum,microtype}
\usepackage[table, xcdraw,dvipsnames]{xcolor}
\usepackage[backend=biber]{biblatex}
\addbibresource{bib/main.bib}
\usepackage{vub}
\usepackage{listings}
\usepackage{lstautogobble}  % Fix relative indenting
\usepackage{xargs}
\usepackage{float}
\usepackage{tikz}
\usepackage{bytefield}
\usepackage{pgf-umlsd}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}
\usepgfplotslibrary{statistics}
\usepgfplotslibrary{groupplots}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{acronym}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{zi4}            % Nice font
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage[toc,page]{appendix}

\definecolor{bluekeywords}{rgb}{0.13, 0.13, 1}
\definecolor{greencomments}{rgb}{0, 0.5, 0}
\definecolor{redstrings}{rgb}{0.9, 0, 0}
\definecolor{graynumbers}{rgb}{0.5, 0.5, 0.5}

\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
\newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
\newcommandx{\info}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}}
\newcommandx{\improvement}[2][1=]{\todo[linecolor=Plum,backgroundcolor=Plum!25,bordercolor=Plum,#1]{#2}}
\newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}

% \usetikzlibrary{external}
% \tikzexternalize[prefix=tikz/]
% \usepackage[a4paper,landscape,hmargin={1cm,1cm}]{geometry}
\usepackage{tikz-timing}[2014/10/29]
\usetikztiminglibrary[rising arrows]{clockarrows}
\usetikzlibrary{fit}
\usetikzlibrary{calc}
\usetikzlibrary{quotes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows,automata}
\usetikzlibrary{backgrounds,calc,shadings,shadows}
\usepackage{circuitikz}
\usetikzlibrary{shapes,decorations.pathreplacing}

\tikzstyle{bag} = [align=center]

\makeatletter
\pgfkeys{/pgf/.cd,
  parallelepiped offset x/.initial=2mm,
  parallelepiped offset y/.initial=2mm
}
\pgfdeclareshape{parallelepiped}
{
  \inheritsavedanchors[from=rectangle] % this is nearly a rectangle
  \inheritanchorborder[from=rectangle]
  \inheritanchor[from=rectangle]{north}
  \inheritanchor[from=rectangle]{north west}
  \inheritanchor[from=rectangle]{north east}
  \inheritanchor[from=rectangle]{center}
  \inheritanchor[from=rectangle]{west}
  \inheritanchor[from=rectangle]{east}
  \inheritanchor[from=rectangle]{mid}
  \inheritanchor[from=rectangle]{mid west}
  \inheritanchor[from=rectangle]{mid east}
  \inheritanchor[from=rectangle]{base}
  \inheritanchor[from=rectangle]{base west}
  \inheritanchor[from=rectangle]{base east}
  \inheritanchor[from=rectangle]{south}
  \inheritanchor[from=rectangle]{south west}
  \inheritanchor[from=rectangle]{south east}
  \backgroundpath{
    % store lower right in xa/ya and upper right in xb/yb
    \southwest \pgf@xa=\pgf@x \pgf@ya=\pgf@y
    \northeast \pgf@xb=\pgf@x \pgf@yb=\pgf@y
    \pgfmathsetlength\pgfutil@tempdima{\pgfkeysvalueof{/pgf/parallelepiped
      offset x}}
    \pgfmathsetlength\pgfutil@tempdimb{\pgfkeysvalueof{/pgf/parallelepiped
      offset y}}
    \def\ppd@offset{\pgfpoint{\pgfutil@tempdima}{\pgfutil@tempdimb}}
    \pgfpathmoveto{\pgfqpoint{\pgf@xa}{\pgf@ya}}
    \pgfpathlineto{\pgfqpoint{\pgf@xb}{\pgf@ya}}
    \pgfpathlineto{\pgfqpoint{\pgf@xb}{\pgf@yb}}
    \pgfpathlineto{\pgfqpoint{\pgf@xa}{\pgf@yb}}
    \pgfpathclose
    \pgfpathmoveto{\pgfqpoint{\pgf@xb}{\pgf@ya}}
    \pgfpathlineto{\pgfpointadd{\pgfpoint{\pgf@xb}{\pgf@ya}}{\ppd@offset}}
    \pgfpathlineto{\pgfpointadd{\pgfpoint{\pgf@xb}{\pgf@yb}}{\ppd@offset}}
    \pgfpathlineto{\pgfpointadd{\pgfpoint{\pgf@xa}{\pgf@yb}}{\ppd@offset}}
    \pgfpathlineto{\pgfqpoint{\pgf@xa}{\pgf@yb}}
    \pgfpathmoveto{\pgfqpoint{\pgf@xb}{\pgf@yb}}
    \pgfpathlineto{\pgfpointadd{\pgfpoint{\pgf@xb}{\pgf@yb}}{\ppd@offset}}
  }
}
\makeatother

\definecolor{listing-background}{HTML}{F7F7F7}
\definecolor{listing-rule}{HTML}{B3B2B3}
\definecolor{listing-numbers}{HTML}{B3B2B3}
\definecolor{listing-text-color}{HTML}{000000}
\definecolor{listing-keyword}{HTML}{435489}
\definecolor{listing-keyword-2}{HTML}{1284CA} % additional keywords
\definecolor{listing-keyword-3}{HTML}{9137CB} % additional keywords
\definecolor{listing-identifier}{HTML}{435489}
\definecolor{listing-string}{HTML}{00999A}
\definecolor{listing-comment}{HTML}{8E8E8E}

\lstset{
  language         = C++,
  numbers          = left,
  xleftmargin      = 2.7em,
  framexleftmargin = 2.5em,
  backgroundcolor  = \color{listing-background},
  basicstyle       = \color{listing-text-color}\linespread{1.0}\small\ttfamily{},
  breaklines       = true,
  frame            = single,
  framesep         = 0.19em,
  rulecolor        = \color{listing-rule},
  frameround       = ffff,
  tabsize          = 4,
  numberstyle      = \color{listing-numbers},
  aboveskip        = 1.0em,
  belowskip        = 0.1em,
  abovecaptionskip = 0em,
  belowcaptionskip = 1.0em,
  keywordstyle     = {\color{listing-keyword}\bfseries},
  keywordstyle     = {[2]\color{listing-keyword-2}\bfseries},
  keywordstyle     = {[3]\color{listing-keyword-3}\bfseries\itshape},
  sensitive        = true,
  identifierstyle  = \color{listing-identifier},
  commentstyle     = \color{listing-comment},
  stringstyle      = \color{listing-string},
  showstringspaces = false,
  escapeinside     = {/*@}{@*/}, % Allow LaTeX inside these special comments
  literate         =
  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
  {à}{{\`a}}1 {è}{{\'e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
  {€}{{\EUR}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1
  {…}{{\ldots}}1 {≥}{{>=}}1 {≤}{{<=}}1 {„}{{\glqq}}1 {“}{{\grqq}}1
  {”}{{''}}1
}
\lstdefinelanguage{none}{
  identifierstyle=,
  commentstyle=,
  stringstyle=,
  keywordstyle=,
}

\def\lav{lavander!90}
\def\oran{orange!30}

\tikzstyle{motes}=[draw,circle,bottom color= gray,
                  top color= white,minimum width=10pt]
\tikzstyle{gateways}=[draw,circle, left color= orange,minimum width=20pt]

\tikzset{l3 switch/.style={
    minimum width=0.75cm,
    minimum height=0.75cm,
    parallelepiped,fill=switch, draw=white,
    parallelepiped offset x=1.75mm,
    parallelepiped offset y=1.25mm,
    path picture={
      \node[fill=white,
        circle,
        minimum size=6pt,
        inner sep=0pt,
        append after command={
          \pgfextra{
            \foreach \angle in {0,45,...,360}
            \draw[-latex,fill=white] (\tikzlastnode.\angle)--++(\angle:2.25mm);
          }
        }
      ]
       at ([xshift=-0.75mm,yshift=-0.5mm]path picture bounding box.center){};
    }
  },
  ports/.style={
    line width=0.3pt,
    top color=gray!20,
    bottom color=gray!80
  },
  rack switch/.style={
    minimum width=1.25cm,
    minimum height=0.25cm,
    parallelepiped,fill=white, draw,
    parallelepiped offset x=2mm,
    parallelepiped offset y=1.25mm,
    xscale=-1,
    path picture={
      \draw[top color=gray!5,bottom color=gray!40]
      (path picture bounding box.south west) rectangle
      (path picture bounding box.north east);
      \coordinate (A-west) at ([xshift=-0.2cm]path picture bounding box.west);
      \coordinate (A-center) at ($(path picture bounding box.center)!0!(path
        picture bounding box.south)$);
      \foreach \x in {0.275,0.525,0.775}{
        \draw[ports]([yshift=-0.05cm]$(A-west)!\x!(A-center)$)
          rectangle +(0.1,0.05);
        \draw[ports]([yshift=-0.125cm]$(A-west)!\x!(A-center)$)
          rectangle +(0.1,0.05);
       }
      \coordinate (A-east) at (path picture bounding box.east);
      \foreach \x in {0.085,0.21,0.335,0.455,0.635,0.755,0.875,1}{
        \draw[ports]([yshift=-0.1125cm]$(A-east)!\x!(A-center)$)
          rectangle +(0.05,0.1);
      }
    }
  },
  server/.style={
    fill=white, draw,
    minimum width=0.35cm,
    minimum height=0.75cm,
    parallelepiped,
    parallelepiped offset x=3mm,
    parallelepiped offset y=2mm,
    xscale=-1,
    path picture={
      \draw[top color=gray!5,bottom color=gray!40]
      (path picture bounding box.south west) rectangle
      (path picture bounding box.north east);
      \coordinate (A-center) at ($(path picture bounding box.center)!0!(path
        picture bounding box.south)$);
      \coordinate (A-west) at ([xshift=-0.575cm]path picture bounding box.west);
      \draw[ports]([yshift=0.1cm]$(A-west)!0!(A-center)$)
        rectangle +(0.2,0.065);
      \draw[ports]([yshift=0.01cm]$(A-west)!0.085!(A-center)$)
        rectangle +(0.15,0.05);
      \fill[black]([yshift=-0.35cm]$(A-west)!-0.1!(A-center)$)
        rectangle +(0.235,0.0175);
      \fill[black]([yshift=-0.385cm]$(A-west)!-0.1!(A-center)$)
        rectangle +(0.235,0.0175);
      \fill[black]([yshift=-0.42cm]$(A-west)!-0.1!(A-center)$)
        rectangle +(0.235,0.0175);
    }
  },
}

\newcommand{\messdash}[4][0]{
  \stepcounter{seqlevel}
  \path
  (#2)+(0,-\theseqlevel*\unitfactor-0.7*\unitfactor) node (mess from) {};
  \addtocounter{seqlevel}{#1}
  \path
  (#4)+(0,-\theseqlevel*\unitfactor-0.7*\unitfactor) node (mess to) {};
  \draw[->,>=angle 60, dashed] (mess from) -- (mess to) node[midway, above]
  {#3};

  \node (#3 from) at (mess from) {};
  \node (#3 to) at (mess to) {};
}


\title{Enabling~IPv6~multi-hop~networks with~Time Slotted~Channel~Hopping and~long~range~LoRA~radios}
\author{Perale Thomas}
\faculty{Science and Bio-Engineering Sciences}
\promotors{Promotors: Prof. Dr. Ir. Kris Steenhaut, Prof. Dr. Ir. Elisa Gonzalez Boix\newline Supervisor: Roald Van Glabbeek}
\pretitle{Master thesis submitted in partial fulfilment of the requirements for the degree of Master of Science in Applied Sciences and Engineering: Computer Science}
\date{2019 - 2020}

\def\triangleH{27.7mm}

\begin{document}
\maketitle

\pagenumbering{roman}


\clearpage
\vspace*{\fill}
\begin{center}
\begin{minipage}{.6\textwidth}
This master's thesis came about (in part) during the period in which higher
education was subjected to  a  lockdown  and  protective  measures  to  prevent
the  spread  of  the  COVID-19  virus.  The  process  of formatting, data
collection, the research method and/or other scientific work the thesis
involved could therefore not always be carried out in the usual manner. The
reader should bear this context in mind when reading this Master's thesis, and
also in the event that some conclusions are taken on board.
\end{minipage}
\end{center}
\vfill % equivalent to \vspace{\fill}
\clearpage

\newpage

\section*{Abstract}

Long range LoRa-based networks made their entry on the market some years ago.
Those networks consist of low power motes that communicate with powerful base
stations in a star of star topology.
Nodes wake up briefly  to send a packet (and receive the ack)  with base
stations that are always listening on all possible channels.
The technical innovation came from the physical layer, incorporated in cheap
low power radio chips that can trade bandwidth for range.
The physical layer allows several orthogonal channels to be created.
At the link layer, contention on a same channel is ruled by a simple random
access medium access control (MAC) protocol, called LoRaWAN,
that randomizes the moments of  retransmission  when a collision has taken
place and an acknowledgement did not arrive.
Practice showed that insufficient coverage (motes cannot reach any of the
available base stations) together with external interference and the simple
MAC protocol for coping with collisions jeopardizes the performance of such a
network.

To solve the blind spot and congestion issues, a  multi-hop Lora
physical layer (Phy) based network is put forward.
A “Lora multi-hop network” consist of motes based on Lora Phy that help each
other to carry packets from source to destination.
These motes must still sleep as much as possible to save energy but need to
wake up at regular moments to be able to serve as relay station for their neighbors.
A mote can only send/listen on one channel at the same time.
This creates the problem for the mote of knowing when to wake up the radio and
on which channel to be able to communicate with the neighbor.
Time Slotted Channel Hopping (TSCH) is a medium access control protocol that
can offer a solution, because it allows nodes to sleep in some time slots and
to wake up in others to listen while a neighbor will be sending in the same slot.
The schedule will have arranged this “rendez vous”.
TSCH has been introduced in for creating ultra-reliable 802.15.4 (ZIGBEE phy)
short range Internet of things (IoT) networks.
Reliability comes from the channel hopping to create frequency diversity and
from the deterministic medium access control.

Organising multihop not only entails the problem of organizing a suitable
link-layer to enable a power efficient communication between neighbor nodes,
it also entails to quest for an appropriate routing protocol to guide packets
from source to destination.
This means more memory will be needed in the Lora motes to accommodate the
protocol stack.
As a starting point for our work, the Contiki-NG lightweight operating system (OS)
that contains a full-fledged IPv6 protocol stack for short range 802.15.4 Phy,
TSCH based networks is taken.

The challenges of porting TSCH from 802.15.4 Phy consists in organizing the
strict timing needed to keep the notion of time slots as well as coping with
the resource constraintness of the motes to fit the code. Another challenge
is to organize the correct interplay between routing and link layer.
After an in-depth study of the details and particularities of LoRa Phy and TSCH,
as well as the study of related work,
I implemented a radio driver for the RN2483 LoRa radio module and adapt
the current implementation of TSCH for 802.15.4 short range in Contiki-NG to
work with it.
I tested the radio driver as well as the adaptation of the protocol I made, by running
a LoRa multi-hop network based on my changes and demonstrated its power efficiency
by using synchronized time slots and reliability to external interference.

\newpage

\section*{Acknowledgement}

I would first like to thank my supervisor Roald Van Glabbeek that introduced me
to the subject and allowed me to catch up with the missing knowledge I had at
the start of the year.
I am especially grateful for his availability and helpfulness that gave me the
right tools to achieve this work.
I would also like to thanks my promotors Prof. Dr. Ir. Kris Steenhaut and Prof.
Dr. Ir. Elisa Gonzalez Boix that provided me with this great subject and helped me
during the year.

% I would also like to thanks the Urlab hackerspace that sparked my interest for
% embedded devices and low power communication protocols over the years.

\newpage

\tableofcontents

\newpage

\listoffigures

\newpage

\section*{List of acronyms}

\input{acro_list}

\newpage

\pagenumbering{arabic}

\input{thesis.tex/chapters/1/index.tex}

\input{thesis.tex/chapters/context/index.tex}

\input{thesis.tex/chapters/driver/index.tex}

\input{thesis.tex/chapters/tsch/index.tex}

\chapter{Conclusion and further work\label{section:conclusion}}

The goal of the thesis aimed to adapt the TSCH implementation of Contiki-NG
running on a Zolertia RE-Mote REV-B to work with the RN2483 LoRa radio,
and thus achieving power-efficient, long range, and reliable IPv6 enabled
multi-hop networks with LoRa radio.

\paragraph{}

After implementing a working driver for the RN2483 LoRa radio module to use
in Contiki, I demonstrated that, by adapting the TSCH MAC protocol to work
LoRa radio, LoRa multi-hops networks are feasible with TSCH and showed:

\begin{itemize}
  \item The network range is enhanced, by routing packets through multiple hops to reach
    its destination.
  \item The network is power efficient, the participants of the network remain synchronized
    to achieve low power communication.
  \item The network is reliable and is resilient to external, by running the network
    over a jammed channel and achieving complete transmission.
\end{itemize}


\paragraph{}

The evaluation of the reliability of TSCH with LoRa was limited by
the drifts issues I got.
I found out those were created by the RN2483 module that did not include any way
to get the precise packet timestamp at reception.
Also, measuring the power efficiency of the network turned out to be meaningless
because of the design of the RN2483 module.

I am confident that measuring those two parameters and achieving a low power
multi-hop network with a radio module like the SX1272 is possible.
This thesis explained what it takes to adapt TSCH for a new LoRa radio
module and the possible re-implementation of this project on a new radio module
will then be facilitated.

\paragraph{}

It can be concluded that the objective of the thesis has only been partially
reached because inadequate hardware was used.
Porting TSCH, an already well studied MAC protocol,
and prooving it can work with the existing multi-hop routing solutions,
represents a step forward in the LoRa multi-hop networking research.
But the lack of reliable measurement on the reliability and the power
efficiency of the network, as well as finding a solution to use all the SF
available shows that further research focussing on those matters should
be conducted.

\section*{Further work\label{section:further}}

The first thing future work need to focus on, is changing the radio platform
used for the tests.
I am confident that achieving a low power multi-hop network with a radio module
like the SX1272 is possible.
This thesis explained what it takes to adapt TSCH for a new LoRa radio
module and the possible re-implementation of this project on a new radio module
Changing the radio driver would create a stable testing device that would allow
us to measure meaningful data for the analysis of power consumption and
reliability of the network.
The following tasks could be conducted on this stable testing device

\begin{itemize}
  \item Determining the optimal TSCH parameters for LoRa.
  \item Finding the optimal period between EB packets to keep a balance between
    low power usage and maintaining synchronization.
  \item Compare how it performs in regards to LoRaWAN. Is the TSCH overhead
    worth it ?
  \item The network capacity and transmission delay.
  \item Running a proper benchmark of the radio usage.
\end{itemize}

The second challenge for the future is to achieve longer distance transmission.
The ideal implementation would adapt its SF depending on the distance between two hops.
Even though the different SF are supported in my implementation, only a single SF is used at the same time.
A lot of LoRa features are lost by using a single SF.
Supporting multiple SF at the same time would increase the network throughput by
allowing parralel communications and increase the transmission range for nodes
with low reception.
The future work on the subject should focus on modifying the protocol to use
different SF.

\paragraph{}

Futur work should conduct a real-world with properly spaced motes to test
the network.

\newpage

\printbibliography

\newpage

\begin{appendices}
\input{thesis.tex/chapters/appendix/index.tex}
\end{appendices}

\end{document}
